Revealing transient strain in geodetic data with Gaussian process regression

### 2017-10-26 RevisedSubmission/hines_hetland_2017_gji.tex
did more

We consider the noise vector, $\mitbf{\eta}_i$, to be composed of a
temporally correlated Gaussian process, $\mathcal{L}_i \sim
\mathcal{GP}(0,C_{\mathcal{L}_i})$, and a vector of uncorrelated
Gaussian noise, $\mitbf{w}_i$, so that
\begin{equation}
\mitbf{\eta}_i = \mathcal{L}_i(\points) + \mitbf{w}_i.
\end{equation}
The standard deviations for $\mitbf{w}_i$ are taken to be the
uncertainties derived for the GNSS displacement solutions,
$\mitbf{\sigma}_i$. The noise vector then has zero mean and the
covariance matrix
\begin{equation}
\mitbf{C}_{\eta_i} = C_{\mathcal{L}_i}(\points,\points) +
                     \mathrm{diag}(\mitbf{\sigma}_i^2).
\end{equation}

@@

The temporally correlated noise in GNSS data has been thoroughly
studied over the past two decades \citep[e.g.,][]{Zhang1997, Mao1999,
Williams2004, Langbein2008}. In these studies, temporally correlated
noise tends to be described with some combination of Brownian motion
(also known as random walk noise or a Weiner process), a first-order
Gauss-Markov (FOGM) process, and/or flicker noise. There is some
physical justification for using Brownian motion as a noise model
because it accurately describes the power spectrum of motion resulting
from instability in geodetic monuments \citep[e.g.,][]{Wyatt1982,
Wyatt1989}. In particular, the power spectrums for Brownian motion and
monument motion are both proportional to $f^{-2}$, where $f$ is
frequency. Brownian motion, by definition, contains a reference time
at which the process begins. Since there is no notion of when noise
``begins" in GNSS data, we do not find Brownian motion to be an
appropriate noise model. On the other hand, a FOGM process also has a
power spectrum that is proportional to $f^{-2}$, above some corner
frequency, and it has no reference time (i.e., a FOGM process is
stationary). We then let $\mathcal{L}_i$ be a spatially uncorrelated
FOGM process, which has the covariance function
\begin{equation}\label{eq:FOGM}
C_{\mathcal{L}_i}\left((\pos,t),(\pos',t')\right) = 
\frac{\beta^2}{2\alpha}
\exp\left(-\alpha|t - t'|\right)\delta_{\pos,\pos'}
\end{equation}
\citep[sec. B2.1]{Rasmussen2006}, where $\delta_{\pos,\pos'}$ is 1 if
$\pos = \pos'$ and 0 otherwise.

@@

By choosing a FOGM process for $\mathcal{L}_i$, we have introduced two
parameters that need to be constrained, $\alpha$ and $\beta$. We
constrain these parameters, which we collectively refer to as
$\mitbf{\theta}$, with the Restricted Maximum Likelihood (REML) method
\citep{Harville1974}. The REML method is conceptually similar to the
Maximum Likelihood Estimation (MLE) method from \citet{Langbein1997},
where we pick $\mitbf{\theta}$ to maximize the likelihood of drawing
the observed data, $\mitbf{d}_i^*$, from the random vector
$\mitbf{d}_i$. However, unlike the MLE method, the REML method
produces unbiased estimates of $\mitbf{\theta}$ \citep[sec.
2.6]{Cressie1992}. We use the REML method to independently estimate
$\mitbf{\theta}$ at 38 continuous GNSS stations in the Pacific
Northwest that are east of 121$^\circ$W. These stations are
sufficiently far from the subduction zone that they are unlikely to
contain transient signal associated with SSEs. We assume that the
noise at these inland stations is representative of the noise at all
the stations considered in this study, which is probably a poor
assumption since the inland stations are subject to distinctly
different climatic conditions. Nonetheless, we find the
$\mitbf{\theta}$ for each inland station and for each displacement
component that maximizes the REML likelihood function
\begin{equation}\label{eq:REML}
\mathcal{L}(\mitbf{\theta}) = \left(\frac{\left|\G^T\G\right|}{(2\pi)^{n-6m} 
                                    \left| \mitbf{C}_{\eta_i}(\mitbf{\theta}) \right| 
                                    \left| \G^T\mitbf{C}_{\eta_i}(\mitbf{\theta})^{-1}\G \right|}\right)^{\frac{1}{2}} 
                              e^{-\tfrac{1}{2}\data_i^{*T}\mitbf{K}(\mitbf{\theta})\data_i^*},
\end{equation}
with respect to $\mitbf{\theta}$, where
\begin{equation}
\mitbf{K}(\mitbf{\theta}) = \mitbf{C}_{\eta_i}(\mitbf{\theta})^{-1} - 
                            \mitbf{C}_{\eta_i}(\mitbf{\theta})^{-1}\G
                            \left(\G^T\mitbf{C}_{\eta_i}(\mitbf{\theta})^{-1}\G\right)^{-1}
                            \G^T\mitbf{C}_{\eta_i}(\mitbf{\theta})^{-1}.
\end{equation}

@@

The distribution of inferred $\alpha$ and $\beta$ are shown in Figure
\ref{fig:NoiseParams}. The estimates of $\beta$ for the easting and
northing components are notably low and are clustered around 0.5
mm/yr$^{0.5}$. The corresponding estimates of $\alpha$ tend to cluster
around 0 yr$^{-1}$. We also estimate $\alpha$ and $\beta$ for the vertical
component of displacements, with the hope that vertical deformation
gradients could reveal some geophysical signal. For the vertical
component, $\beta$ is relatively large with a median value of 13.5
mm/yr$^{0.5}$. The inferred values for $\alpha$ are also higher for
the vertical component with a median value of 8.21 yr$^{-1}$. In
Figure \ref{fig:NoiseSamples}, we use the median values of $\alpha$
and $\beta$ to generate two realizations of FOGM noise for each
component. The realizations span five years, and the easting and
northing realizations drift by about 1 mm over these five years. In
the context of detecting SSEs, which produce several mm's of surface
displacement on the time-scale of weeks, the estimated FOGM noise for
the easting and northing component is negligible. In contrast, the
estimated FOGM noise for the vertical component is larger than the
signal we would expect from SSEs. We suspect that the higher amplitude
for the FOGM noise in the vertical component is accommodating for
deficiencies in our rather simple seasonal model. Based on this
analysis, we henceforth ignore temporally correlated noise in the
easting and northing component because of its low amplitude. We also
do not use vertical displacements because of the presumably low
signal-to-noise ratio.

>>>>

We consider the noise vector, $\mitbf{\eta}_i$, to be composed of a
temporally correlated Gaussian process, $z_i \sim
\mathcal{GP}(0,C_{z_i})$, and a vector of uncorrelated
Gaussian noise, $\mitbf{w}_i$, so that
\begin{equation}
\mitbf{\eta}_i = z_i(\points) + \mitbf{w}_i.
\end{equation}
The standard deviations for $\mitbf{w}_i$ are taken to be the
uncertainties derived for the GNSS displacement solutions,
$\mitbf{\sigma}_i$. The noise vector then has zero mean and the
covariance matrix
\begin{equation}
\mitbf{C}_{\eta_i} = C_{z_i}(\points,\points) +
                     \mathrm{diag}(\mitbf{\sigma}_i^2).
\end{equation}

@@

The temporally correlated noise in GNSS data has been thoroughly
studied over the past two decades \citep[e.g.,][]{Zhang1997, Mao1999,
Williams2004, Langbein2008}. In these studies, temporally correlated
noise tends to be described with some combination of Brownian motion
(also known as random walk noise or a Weiner process), a first-order
Gauss-Markov (FOGM) process, and/or flicker noise. There is some
physical justification for using Brownian motion as a noise model
because it accurately describes the power spectrum of motion resulting
from instability in a geodetic monument \citep[e.g.,][]{Wyatt1982,
Wyatt1989}. In particular, the power spectrums for Brownian motion and
monument motion both decay proportionally to $f^{-2}$, where $f$ is
frequency. However, Brownian motion necessarily contains a reference
time at which the process begins. Since there is no notion of when
noise ``begins" in GNSS data, we do not find Brownian motion to be an
appropriate noise model. On the other hand, a FOGM process has no
reference time (i.e., it is stationary), and its power spectrum,
\begin{equation}\label{eq:FOGMPS}
S(f) = \frac{\beta^2}{(2 \pi f)^2 + \alpha^2},
\end{equation}
decays proportionally to $f^{-2}$ above the cutoff frequency $\alpha /
(2 \pi)$. We then choose $z_i$ to be a spatially
uncorrelated FOGM process, which has the covariance function
\begin{equation}\label{eq:FOGM}
C_{z_i}\left((\pos,t),(\pos',t')\right) = 
\frac{\beta^2}{2\alpha}
\exp\left(-\alpha|t - t'|\right)\delta_{\pos,\pos'},
\end{equation}
where $\delta_{\pos,\pos'}$ is 1 if $\pos = \pos'$ and 0 otherwise.
See \citet[sec. B2]{Rasmussen2006} for more information on the FOGM
process.

@@

By choosing a FOGM process for $z_i$, we have introduced two
parameters that need to be constrained, $\alpha$ and $\beta$. We
constrain these parameters, which we collectively refer to as
$\mitbf{\theta}$, with the Restricted Maximum Likelihood (REML) method
\citep{Harville1974}. The REML method is conceptually similar to the
Maximum Likelihood Estimation (MLE) method from \citet{Langbein1997},
in that we attempt to pick $\mitbf{\theta}$ to maximize the
probability of drawing the observed data, $\mitbf{d}_i^*$, from the
random vector $\mitbf{d}_i$. However, unlike the MLE method, the REML
method produces unbiased estimates of $\mitbf{\theta}$ \citep[sec.
2.6]{Cressie1992}. We use the REML method to estimate $\mitbf{\theta}$
at 38 continuous GNSS stations in the Pacific Northwest that are east
of 121$^\circ$W. These stations are sufficiently far from the
subduction zone that they are unlikely to contain transient signal
associated with SSEs, allowing us to ignore displacements from $u_i$
in $\mitbf{d}_i$. We assume the noise at these inland stations is
representative of the noise at all the stations considered in this
study, which is probably a poor assumption since the inland stations
are subject to distinctly different climatic conditions. Nonetheless,
we find $\mitbf{\theta}$ for each of the inland stations and for each
displacement component by maximizing the REML likelihood function

\begin{equation}\label{eq:REML}
\mathcal{L}(\mitbf{\theta}) = \left(\frac{\left|\G^T\G\right|}{(2\pi)^{n-6m} 
                                    \left| \mitbf{C}_{\eta_i}(\mitbf{\theta}) \right| 
                                    \left| \G^T\mitbf{C}_{\eta_i}(\mitbf{\theta})^{-1}\G \right|}\right)^{\frac{1}{2}} 
                              e^{-\tfrac{1}{2}\data_i^{*T}\mitbf{K}(\mitbf{\theta})\data_i^*}
\end{equation}
with respect to $\mitbf{\theta}$, where
\begin{equation}
\mitbf{K}(\mitbf{\theta}) = \mitbf{C}_{\eta_i}(\mitbf{\theta})^{-1} - 
                            \mitbf{C}_{\eta_i}(\mitbf{\theta})^{-1}\G
                            \left(\G^T\mitbf{C}_{\eta_i}(\mitbf{\theta})^{-1}\G\right)^{-1}
                            \G^T\mitbf{C}_{\eta_i}(\mitbf{\theta})^{-1}.
\end{equation}

@@

The distribution of inferred $\alpha$ and $\beta$ are shown in Figure
\ref{fig:NoiseParams}. The estimates of $\beta$ for the easting and
northing components are clustered around 0.5 mm/yr$^{0.5}$. The
corresponding estimates of $\alpha$ tend to cluster around 0
yr$^{-1}$, indicating that the power spectrum of noise indeed decays
proportionally to $f^{-2}$ over a wide band of frequencies. We also
estimate $\alpha$ and $\beta$ for the vertical component of
displacements, with the hope that vertical deformation gradients could
reveal some geophysical signal. For the vertical component, $\beta$ is
relatively large with a median value of 13.5 mm/yr$^{0.5}$. The
inferred values for $\alpha$ are also higher for the vertical
component with a median value of 8.21 yr$^{-1}$. In Figure
\ref{fig:NoiseSamples}, we use the median values of $\alpha$ and
$\beta$ to generate two realizations of FOGM noise for each component.
The realizations span five years, and the easting and northing
realizations drift by about 1 mm over these five years. In the context
of detecting SSEs, which produce several mm's of surface displacement
on the time-scale of weeks, the estimated FOGM noise for the easting
and northing component is negligible. In contrast, the estimated FOGM
noise for the vertical component is larger than the signal we would
expect from SSEs. We suspect that the higher amplitude for the FOGM
noise in the vertical component is accommodating for deficiencies in
our rather simple seasonal model. Based on this analysis, we
henceforth ignore temporally correlated noise in the easting and
northing component because of its low amplitude. We also do not use
vertical displacements because of the presumably low signal-to-noise
ratio.
